---
title: "Coursework Project"
author: "Student N°200580234"
---

#### Loding relevant librarys

```{r}
library("dplyr")
library("tidyr")
library("tidyverse")
library("ggplot2")
library('vioplot')
library('igraph')
library('reshape2')
library('mlr3')
library('mlr3learners')
library('mlr3pipelines')
library('xgboost')
```

#### Importing, cleaning, preparing data
```{r}
#Import Data
planes_data <- read.csv("plane-data.csv")
airport_data <- read.csv("airports.csv")
flights <- read.csv("sampled_data.csv")
```

#### Dropping cancelled flights
```{r}
flights <- filter(flights, Cancelled == 0)
summary(flights)
```
#### Clasifying delays and previous aircraft delay
```{r}
flights <- flights %>% 
  mutate(delayed = case_when(flights$ArrDelay >= 15 ~ 1, TRUE ~ 0)) %>% mutate(predelayed = case_when(flights$LateAircraftDelay >= 15 ~ 1, TRUE ~ 0))
tail(flights)
```
## Question 1
#### Selecting relevant columns and dropping NAs
```{r}
delay_columns <- c("Year","Month","DayofMonth","DayOfWeek","ArrDelay","CRSDepTime","delayed")
delays_df <- select(flights, delay_columns)
delays_df <- delays_df %>% drop_na()
head(delays_df)
```

#### Calculating Ratio per day of week
```{r}
delays_dayofweek <- delays_df %>% 
  group_by(DayOfWeek) %>% 
  summarise(delayes = sum(delayed), total_flight = n(), flight_ratio= delayes/total_flight) %>%
  arrange(DayOfWeek)

head(delays_dayofweek)
  
```

#### Plotting delay per day of week
```{r} 
gr_dayofweek <- ggplot(delays_dayofweek, aes(x= DayOfWeek, y= flight_ratio, colour= DayOfWeek, fill = DayOfWeek)) + 
  geom_bar(stat = 'identity') +
  labs(title = "Delay per day of Week", x = "Day of Week", y = "Daily ratio")
  
gr_dayofweek
```

```{r} 
vioplot(DayOfWeek ~ delayed, data = delays_df)
```
#### Calculating ratio per month
```{r}
delays_month <- delays_df %>% 
  group_by(Month) %>% 
  summarise(delayes = sum(delayed), total_flight = n(), flight_ratio= delayes/total_flight) %>%
  arrange((Month))
head(delays_month)
```
#### Plotting ratio per month
```{r}
gr_month <- ggplot(delays_month, aes(x= Month, y= flight_ratio, colour= Month, fill = Month)) + 
  geom_bar(stat = 'identity') +
  labs(title = "Delay per Month", x = "Month", y = "Monthly ratio")

gr_month
```

```{r} 
vioplot(Month ~ delayed, data = delays_df)
```
#### Calculating ratio per hour
```{r}
delays_time <- delays_df %>% mutate(Time = round(CRSDepTime/100,0))

delays_time <- delays_time %>% 
  group_by(Time) %>% 
  summarise(delayes = sum(delayed), total_flight = n(), flight_ratio= delayes/total_flight) %>%
  filter(Time >= 5) %>%
  arrange(Time) 
head(delays_time)
```
#### Plotting ratio per hour
```{r}
gr_time <- ggplot(delays_time, aes(x= Time, y= flight_ratio, colour= Time, fill = Time)) + 
  geom_bar(stat = 'identity') +
  labs(title = "Delay per Hour", x = "Hour of day", y = "Hourly ratio")

gr_time
```

```{r} 
vioplot(CRSDepTime ~ delayed, data = delays_df)
```

## Question 2
#### Joining Dataframes and wrangling data

```{r} 
flights_planes <- flights %>%  
  left_join(planes_data, by= c("TailNum" = "tailnum"))
flights_age_columns = c("Year","Month","DayofMonth","delayed","year")
flights_planes <- select(flights_planes,flights_age_columns)
flights_planes  <- flights_planes %>% 
  drop_na("year")
flights_planes[,5] <- sapply(flights_planes[,5], as.numeric)
flights_planes <- drop_na(flights_planes)
flights_planes <- flights_planes %>% filter( year > 0 & year<2007)
flights_planes <- flights_planes %>% mutate(Age = Year - year)
head(flights_planes)
```
#### Plotting delays per age
```{r}
delays_age <- flights_planes %>% 
  group_by(Age) %>%
  summarise(delayes = sum(delayed), total_flight = n(), flight_ratio= delayes/total_flight) %>%
  arrange(Age)

gr_age <- ggplot(delays_age, aes(x= Age, y=flight_ratio)) + 
  geom_histogram(stat = 'identity')
gr_age
```

```{r} 
vioplot(Age ~ delayed, data = flights_planes)
```

## Question 3
#### Selecting Columns
```{r} 
flights_people <- flights[c('Year','Month','Origin','Dest')]
head(flights_people)
```

#### Data Wranlging and Joining data
```{r}
flights_people <- flights_people %>%
  left_join(airport_data, by = c("Origin" = "iata"))
flights_people <- rename(flights_people,  "Origin_city" = "city")
flights_people <- flights_people %>% left_join(airport_data, by = c("Dest" = "iata"))
flights_people <- rename(flights_people, "Destination_city" = 'city')
flights_people_columns <- c("Year","Month", "Origin_city", "Destination_city")
flights_people <- select(flights_people,flights_people_columns)
flights_people <- drop_na(flights_people)
head(flights_people)
```
#### Calculating ratio per Origin city and selecting top
```{r}
flights_people_o <- flights_people %>%
  group_by(Origin_city, Year) %>% 
  summarise(Count = n()) %>% 
  dcast(formula = Origin_city ~ Year,
        value.var = 'Count' )
flights_people_o <- flights_people_o %>%
  mutate(Difference = abs(flights_people_o[,3] - flights_people_o[,2]), Total = (flights_people_o[,3] + flights_people_o[,2]), Ratio = Difference*100 /Total )
flights_people_o <- flights_people_o %>% 
  filter(Total > 60) %>%
  arrange(desc(Ratio))
 

head(flights_people_o)
```
#### Plotting differences in frequency per year
```{r}
flights_people_o <- flights_people_o %>%
  top_n(5)
flights_people_o <- flights_people_o[c(1,2,3)]
flights_people_o <- melt(flights_people_o, id.vars = "Origin_city", variable.name = "Year", value.name = "Count")


gr_origin <- ggplot(flights_people_o, aes(fill = Year, x= Origin_city, y = Count)) + 
  geom_bar(position = "dodge", stat = 'identity')
gr_origin
```
#### Calculating ratio per Destination city and selecting top
```{r}
flights_people_d <- flights_people %>%
  group_by(Destination_city, Year) %>% 
  summarise(Count = n()) %>% 
  dcast(formula = Destination_city ~ Year,
        value.var = 'Count' )
flights_people_d <- flights_people_d %>%
  mutate(Difference = abs(flights_people_d[,3] - flights_people_d[,2]), Total = (flights_people_d[,3] + flights_people_d[,2]), Ratio = Difference*100 /Total )
flights_people_d <- flights_people_d %>% 
  filter(Total > 60) %>%
  arrange(desc(Ratio))
 

head(flights_people_d)
```
#### Plotting difference for Destination
```{r}
flights_people_d <- flights_people_d %>%
  top_n(5)
flights_people_d <- flights_people_d[c(1,2,3)]
flights_people_d <- melt(flights_people_d, id.vars = "Destination_city", variable.name = "Year", value.name = "Count")


gr_destination <- ggplot(flights_people_d, aes(fill = Year, x= Destination_city, y = Count)) + 
  geom_bar(position = "dodge", stat = 'identity')
gr_destination

```
#### Grouping by origin and destination and selecting top with most frequency for year 2005
```{r}
flights_people_2005 <- filter(flights_people, Year == 2005)
flights_people_2005 <- flights_people_2005 %>% 
  group_by(Origin_city,Destination_city) %>%
  summarise(Total = n()) %>%
  arrange(desc(Total))
  
flights_people_2005 <- flights_people_2005[1:60,]
flights_people_2005
```
#### Creating list of nodes filtered by data from previous step
```{r}
airport_data_cities <- airport_data[,c(3,1,2,4,5,6,7)]
airport_data_cities <- airport_data_cities %>% distinct(city) %>% drop_na(city)
airport_data_cities_2005 <- airport_data_cities %>% filter(city %in% flights_people_2005$Origin_city | city %in% flights_people_2005$Destination_city ) 
airport_data_cities_2005
```
#### Plotting network visualization for year 2005
```{r}
network_2005 <- graph_from_data_frame(d=flights_people_2005, vertices = airport_data_cities_2005, directed = T)
network_2005 <- simplify(network_2005, remove.multiple = F, remove.loops = T)
plot(network_2005, edge.arrow.size = 1)
```
#### Grouping by origin and destination and selecting top with most frequency for year 2006
```{r}
flights_people_2006 <- filter(flights_people, Year == 2006)
flights_people_2006 <- flights_people_2006 %>% 
  group_by(Origin_city,Destination_city) %>%
  summarise(Total = n()) %>%
  arrange(desc(Total))
  
flights_people_2006 <- flights_people_2006[1:60,]
flights_people_2006
```
#### Creating list of nodes filtered by data from previous step
```{r}
airport_data_cities_2006 <- airport_data_cities %>% filter(city %in% flights_people_2006$Origin_city | city %in% flights_people_2006$Destination_city ) 
airport_data_cities_2006
```
#### Plotting network visualization for year 2006
```{r}
network_2006 <- graph_from_data_frame(d=flights_people_2006, vertices = airport_data_cities_2006, directed = T)
network_2006 <- simplify(network_2006, remove.multiple = F, remove.loops = T)
plot(network_2006, edge.arrow.size = 1)
```
## Question 4
#### Selecting delayed flights
```{r}
flights_Late <- filter(flights, delayed == 1)
head(flights_Late)
```
#### Calculating percentage of delayed flights with previous aircraft delay
```{r}
sum(flights_Late$predelayed)*100/sum(flights_Late$delayed)
```
#### Selecting flights with previous delayed aircraft
```{r}
flights_pre <- filter(flights, predelayed == 1)
head(flights_pre)
```
#### Calculating percentage of flights with previous aircraft delay that where also delayed
```{r}
sum(flights_pre$delayed)*100/sum(flights_pre$predelayed)
```
#### Looking for airport with most delays in a day by origin
```{r}
worst_day_o <- flights %>%
  group_by(Year, Month, DayofMonth, Origin) %>%
  summarise(Sum = sum(delayed))%>%
  arrange(desc(Sum))
worst_day_o <- worst_day_o[1:5,]
worst_day_o
```
#### Inspecting data for that day
```{r}
dep_worst_o <- filter(flights, Year == 2006 & Month == 10 & DayofMonth == 17 & Origin == 'ATL')
dep_worst_o
```
#### Looking for airport with most delays in a day by destination
```{r}
worst_day_d <- flights %>%
  group_by(Year, Month, DayofMonth, Dest) %>%
  summarise(Sum = sum(delayed))%>%
  arrange(desc(Sum),Year)
worst_day_d <- worst_day_d[2,]
worst_day_d
```
#### Inspecting that day
```{r}
dep_worst_d <- filter(flights, Year == 2006 & Month == 10 & DayofMonth == 19 & Origin == 'ATL')
dep_worst_d
```
## Question 5
#### Setting as factor relevant categorical variables
```{r}
flights$delayed <- factor(flights$delayed)
flights$Dest <- factor(flights$Dest)
flights$Origin <- factor(flights$Origin)
flights$TailNum <- factor(flights$TailNum)
flights$UniqueCarrier <- factor(flights$UniqueCarrier)
```
#### Dropping NAs and sampling data because of performance issues with my PC
```{r}
flights_model <- flights %>% drop_na()
flights_model <- sample_n(flights_model,500)
```

#### Creating task for model
```{r}
task <- TaskClassif$new('flights_model', backend = flights_model, target = "delayed")
task$select(c("Year", "Month","DayOfWeek","CRSDepTime","UniqueCarrier","Origin","Dest","TailNum","Distance","LateAircraftDelay"))
measure <- msr("classif.ce")
task
```
#### Creating encoder for categorical variables
```{r}
fencoder <- po("encode", method = "treatment", affect_columns = selector_type("factor"))
```

#### Splitting data into test and train
```{r}
set.seed(1)
train_set <- sample(task$nrow, 0.8*task$nrow)
test_set <- setdiff(seq_len(task$nrow), train_set)
```

#### Dropping levels in categorial variable
```{r}
levels(droplevels(flights_model$Origin))
```
#### Fitting logistic regression
```{r}
learner_lr <- lrn("classif.log_reg")
gc_lr <- po("scale") %>>% fencoder %>>% po("imputemean") %>>% po(learner_gr)
glrn_lr <- GraphLearner$new(gc_lr)

glrn_lr$train(task, row_ids = train_set)
glrn_lr$predict(task, row_ids = test_set)$score()
```
#### Fitting Gradient Boosting
```{r}
learner_gr <- lrn("classif.xgboost")
gc_gr <-po("scale") %>>% fencoder %>>% po("imputemean") %>>% po(learner_gr)
glrn_gr <- GraphLearner$new(gc_gr)
glrn_gr$train(task, row_ids = train_set)
glrn_gr$predict(task, row_ids = test_set)$score()
```















